\thetitle

### 2021-02-11 writing/reviews/pldi-rebuttal-final.md

> It's interesting to note that there seems to be no convincing analogous
> experimental or modelling-based justification for plain old static
> typing, i.e. showing that it somehow presents a net gain to the
> programmer. Rather, this has simply been posited/assumed by a very long
> line of work. I am not defending that state of affairs, but it points to
> the difficulty of showing conclusively that something truly helps
> programmers. I can see value in this sort of simulation-style approach,
> and it is no less convincing than user studies. So I'd be interested to
> hear from the author(s) if they have any more arguments that (1) I've
> underestimated the results' conclusiveness, or (2) this is a
> novel/interesting family of methods that might be pursued more widely,
> or (3) that deeper experiments building on these ideas might yield more
> compelling insights.

For points 1 and 3 please see our discussion at the beginning of the
reply. For point 2, the idea of the rational programmer is a new one so
far applied only to blame in contracts (Lazarek et al. POPL 2020) and blame in
gradual typing (our work). We do believe it can have a similar effect in
programming languages as homo economicus in economics.

That said, as posed, this comment seems to express a significant
mis-characterization of the submission. It is not the benefits of a
gradual type system that are in doubt. The experimental setup keeps the
type system constant but allows to answer the question which of several
checking and blaming regimes (for enforcing types) provides the best
explanatory messages in case of violations. Hence the analogous question
for purely static type systems would ask which of several reporting
schemes provides the best explanatory message for type errors. In the case
of simply typed languages and even languages with local type inference,
this question is basically meaningless. In the case of languages with HM
type inference, the question has been implicitly raised for four decades
with the development of alternative ways of finding the source of
inference conflicts. One recent example is SHErrLoc (Zhang et al. TOPLAS
2017).

If the reviewer is indeed interested in the question of whether type systems
help programmers --- a question that this submission does _not_ ask --- then
Hanenberg and Stefik's papers are a good starting point (e.g. ICSE 2014).

@@

> line 34: "then their compilers remove types and rely on the built-in
> safety checks of the underlying language to catch any problems". This
> reads oddly in context. Didn't they just do a bunch of static checking?
> So they are *not* relying just on dynamic checks to catch problems?
> Maybe there is something more accurate to say here... e.g. no dynamic
> check is removed, or something like that.

In Erasure, a program can type check and still have type-level errors
because some type annotations may be missing. Section 3 illustrates
exactly this scenario by example. In contrast to Natural and Transient,
types in Erasure do not result in run time checks so the only way to
detect type-level errors that the type checker could not find is the
built-in checks of the language's primitives.

@@

> line 60: around this paragraph the writing started to grate. There's no
> need to generalise about what people do or think, and the "As a matter
> of fact... simply..." style is somewhere between laboured and
> patronising. It's better to more plainly state the gap in the literature
> that you're addressing. Throughout the paper, much space could be saved
> by writing in a more direct style.

Noted, we will rephrase.

> line 69: be explicit that Lazarek et al were (as I later gathered) doing
> something about blame to do with higher-order contracts but not gradual
> types

Thanks, we will add a more detailed comparison along the lines of the
discussion at the beginning of our response.

@@

> line 70: what does it mean to "follow" the slogan? Was unclear to me.

The sentences that follow line 70 are meant to expand on the slogan. We
will adjust the prose to make it clear.

@@

> line 187: what is a "component"? Should be easy to define.

In the context of Typed Racket, a component is a module.

> line 277: what is "responsibility" of a "party", exactly? Are "party",
> "module" and "component" all the same things?

Yes in this context. Party and component refer to the entities that can be
blamed, which in Typed Racket is modules.

@@

> line 293: I was wondering what constitutes a boundary crossing. Clearly,
> passing by function call or return crosses from the caller's module to
> the callee's. What about values exchanged through reads/writes to shared
> state?

Yes, exchange of values via state also constitutes a boundary crossing.
Some of our benchmarks do exactly that. We will clarify what a boundary
crossing is in the prose.

@@

> line 313: it threw me that a program might not fail but "produce a wrong
> result". If it could be caught by the gradual type system, why can't it
> be caught at run time? My best get at explaining this is by what I wrote
> above, i.e. it's a consequence of Typed Racket's more refined static
> notion of type. In certain other systems this wouldn't be possible,
> because the static checker would only catch (albeit earlier) errors that
> would be caught at run time, so there would be no basis to call the
> result "wrong".

No. A buggy program can produce a value in Erasure where in Natural or
Transient it produces a run time type error. Greenman et al. [10] prove
formally this proposition but it is easy to construct an example that
demonstrates the issue. For instance if the client in Figure 1 just
formats the result of 'crypto-pack' to a string and returns it, then the
client will terminate without an error but its string result will be
wrong; it will be a string representation of a hash rather than a list.
The reason is that the `format` primitive of Racket does not care what its
argument is.

@@

> line 424: "fully typed correct programs" -- presumably your method could
> also work with not-yet-fully-typed correct programs, just not ranging
> over the entire lattice in those cases. I was wondering whether that
> might give different/interesting results.

Yes. In fact, we already do this for programs that depend on library code.

> line 517: the difference between #1 and #2 here again relies on the
> surprising property (to the unfamiliar) that Typed Racket has a notion
> of "type-level mistake" that doesn't surface under erasure (as an
> exception or whatever) but also is not a false positive (i.e. you're not
> talking about conservativeness of static checking). Line 538's
> "unavoidable" claim is probably also true only in such a context.

Yes please see our answers to your questions about Erasure, line 34 and
line 313.

@@

> line 756: "checked the value's type.." -- and the check passed!?

Indeed, Transient checks only the top-level type constructor of a
value. We will add a reminder of section 3's discussion of Transient
checks to the prose here.
